---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
First, we load some libraries and the needed the data using `read.csv` and then we make the following adjustments:
1. Convert `date` column to date format
2. Convert the interval to time

```{r echo=TRUE}
#first, load in libraries
library(ggplot2)
library(tibble)
library(lubridate)
library(stringr)
library(dplyr)
library(lattice)

#load data
activityData <- read.csv(file="activity.csv",header=TRUE)

#convert date column to date
activityData$date <- as.Date(as.character(activityData$date),format='%Y-%m-%d')
activityData$interval <- as.factor(activityData$interval)

head(activityData)


```

Now that the data is loaded in and processed, we can proceed to the graphs.


## What is mean total number of steps taken per day?

First, we get the total number of steps and the total number of days.
```{r echo=TRUE, cache=TRUE}
totalSteps <- sum(activityData$steps, na.rm=TRUE)
totalDays <- length(unique(activityData$date))
```
There were `r totalSteps` steps taken over `r totalDays` days.

Next, we find the mean:

```{r echo=TRUE}
averageSteps <- as.integer(totalSteps/totalDays)
```
The mean number of steps taken per day is `r averageSteps` steps.

Let's plot the number of steps per day as a histogram.
```{r echo=TRUE}
daySteps <- aggregate(activityData$steps, by=list(activityData$date),FUN=sum, na.rm=TRUE)

hist(daySteps$x,xlab="Steps per day",main="Histogram of daily step count")
```

## What is the average daily activity pattern?

First, we calculate the average number of steps per interval
```{r echo=TRUE}
stepList <- tapply(activityData$steps, activityData$interval, mean, na.rm=TRUE)
```
This returns a list.

Next, we convert the list into a dataframe and ensure the interval is stored as a column rather than the row names.
```{r echo=TRUE}
stepDf <- as.data.frame(stepList)
stepDf <- rownames_to_column(stepDf, var="interval")
```

Now that we have our data frame, we can turn this into a graph.

```{r echo=TRUE}
stepGraph <- ggplot(data=stepDf, aes(x=as.integer(interval),y=stepList))

stepGraph <- stepGraph +
             geom_line() +
             labs(x="Interval",
                  y="Average Steps",
                  title="Daily Activity Pattern")
stepGraph

```

There is clearly a max average number of steps across all intervals.
```{r echo=TRUE}
maxIndex <- unname(which.max(stepList))
maxTime <- names(stepList[maxIndex])
```
The max average occurs at `r maxTime`.


## Imputing missing values

To impute the missing values, it will be helpful to first know how many missing values there are.
```{r echo=TRUE}
numMissingVals <- sum(is.na(activityData$steps))
```

There are `r numMissingVals` missing values. 

To impute these missing values, we'll replace them with the interval average.
```{r echo=TRUE}
#Create our data set where we'll impute data
imputedData <- activityData

#loop over original data set and impute values for NAs
for(i in 1:length(activityData$steps)){
    if(is.na(activityData$steps[i])){
        intervalVal <- as.character(activityData$interval)[i]
        imputedData$steps[i] <- unname(stepList[intervalVal])
    }
}
```

Now that we've imputed data, lets create a histogram of the total number of steps taken each day.
```{r echo=TRUE}
daySteps <- aggregate(imputedData$steps, by = list(imputedData$date), FUN = sum)

hist(daySteps$x,xlab="Steps per day",main="Histogram of daily step count")
```

Let's also grab an updated mean and median for total steps per day

```{r echo=TRUE}
mnSteps <- as.integer(mean(daySteps$x))
mdnSteps <- as.integer(median(daySteps$x))
```

The mean number of steps per day is `r mnSteps` and the median number of steps per day is `r mdnSteps`. These numbers are clearly higher than what we initially found. This is because we are added additional steps each day with our imputed values.

## Are there differences in activity patterns between weekdays and weekends?

First, lets add a new variable to our dataset that returns whether or not the day is a weekday or not.

```{r echo=TRUE}
activityData <- mutate(activityData, Is.Weekend = ifelse(weekdays(activityData$date) 
                                                            %in% c("Saturday","Sunday"),1,0))
activityData$Is.Weekend <- factor(activityData$Is.Weekend,
                                  levels=c(0,1),
                                  labels=c("Weekday","Weekend"))
```

Next, lets break this out into two plots: one for average steps by interval on weekdays and one for average steps by interval on weekends. We'll also convert the interval variable to an integer to make graphing easier.

```{r echo=TRUE}
activityAvg <- aggregate(activityData$steps, by=list(activityData$interval, activityData$Is.Weekend), FUN=mean, na.rm=TRUE)

activityAvg$Group.1 <- as.integer(levels(activityAvg$Group.1))

xyplot(x ~ Group.1 | Group.2, data=activityAvg, 
       layout=c(1,2),
       type="l",
       ylab="Steps",
       xlab="interval"       ,
       scales=list(y=list(tick.number=5),
                   x=list(tick.number=5)))
```
